---
title: "`cloudfs` package overview"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_width: 7
    fig_height: 7
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Working With Cloud File Storage}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(knitr)
opts_chunk$set(
	comment = "",
	#fig.width = 12, 
	message = FALSE,
	warning = FALSE,
	tidy.opts = list(
		keep.blank.line = TRUE,
		width.cutoff = 150
		),
	options(width = 150),
	eval = FALSE,
	error = TRUE)
```

## Introduction

### Project assets

In terms of digital assets a data analysis project usually can be divided into
three parts:

-   **input data**: primarily tabular data (csv, sav, ...), but not limited to 
    it; can be e.g. raw text

-   **code sources**: R or Rmd scripts where the data processing happens

-   **outputs**: spreadsheets, plots, presentations derived from input data

It's common to segregate the storage of code sources, input data, and outputs.
Code sources are typically housed on platforms like GitHub, optimized for
version control and collaboration.

Storing large input data on such platforms can be inefficient. Systems like git
can become overwhelmed when tracking changes to sizable data files. Hence, cloud
storage solutions like Amazon S3 are commonly chosen for their ability to handle
vast datasets.

Outputs, being the results destined for sharing and review, are usually stored
separately from the code. Platforms like Google Drive are preferred for this
purpose. Their user-friendly interfaces and easy sharing options ensure
stakeholders can readily access and review the results.

### Cloud roots

In this context, we won't differentiate between assets being inputs or outputs.
We'll simply refer to both as **artifacts**, setting them apart from code.

The package includes functions for working with both **S3** and **Google
Drive**. In our operations, we use the term **cloud root** to denote the primary
folder on either platform, which holds the project's artifacts. Typically, the
root folder contains subfolders like "plots", "data", and "results", though the
exact structure isn't crucial. A project can have a root on S3, on Google Drive,
or on both platforms simultaneously.

### Motivation

Consider a typical task: uploading a file to Amazon S3 using the `aws.s3`
package as an illustrative example. Imagine we're attempting to upload an R
model saved as an RDS file located at "models/glm.rds". This file is destined
for the "project-1" directory within the "project-data" bucket on S3,
representing the dedicated S3 root for this project:

```{r eval=FALSE}
aws.s3::put_object(
  bucket = "project-data",
  object = "project-1/models/glm.rds",
  file = "models/glm.rds"
)
```

Note the following:

1. **Location Redundancy**: Given that our project's primary interactions are
with the "project-1" folder in the "project-data" bucket, we're consistently
faced with specifying this static location.

2. **Path Duplication**: Both our local system and S3 use matching paths:
"models/glm.rds". This uniformity is typically more practical than making
exceptions.

Given the repetitive nature of this code, there's room for a more streamlined
approach. This is where the `cloudfs` package comes in. Once set up, uploading
becomes much neater:

```{r eval=FALSE}
cloud_s3_upload("models/glm.rds")
```

## Package walkthrough

### Setting up a root

To begin working with the `cloudfs` package in your R project, first set up a
cloud root. For S3 use `cloud_s3_attach()`, for Google Drive, use the
`cloud_drive_attach()` function. Let's set up a Google Drive root:

```{r eval=FALSE}
library(cloudfs)
cloud_drive_attach()
```

Upon execution, you'll be prompted to input the URL of the intended Google Drive
folder to serve as the project's root. This location is then registered in the
project's DESCRIPTION file. To conveniently access this directory in the future,
execute `cloud_drive_browse()`.

### Types of interactions

Now let's talk about actual interactions with the cloud storage. Data transfer
actions can be categorized by two parameters:

1.  **direction** -- whether you're uploading data to the cloud or retrieving
    data from it.

2.  **file or R object** -- using `cloudfs`, you can not only upload and 
    download files from cloud storages but also directly read from and write
    objects to the cloud.

`cloudfs` functions for moving files use "upload" or "download" in their names.
Functions for direct reading or writing use "read" or "write". S3-specific
functions contain "s3", while Google Drive ones use "drive".

+-------------------------+-------------------------+-------------------------+
|                         | to cloud                | from cloud              |
+=========================+=========================+=========================+
| file                    | `cloud_s3_upload`<br>   | `cloud_s3_download`<br> |
|                         | `cloud_drive_upload`    | `cloud_drive_download`  |
+-------------------------+-------------------------+-------------------------+
| R object                | `cloud_s3_write`<br>    | `cloud_s3_read`<br>     |
|                         | `cloud_drive_write`     | `cloud_drive_read`      |
+-------------------------+-------------------------+-------------------------+


### Practical Examples

Here, we'll demonstrate the hands-on application of `cloudfs` functions for data
transfer.

Upon successfully completing the `cloud_drive_attach()` process, your project
will be associated with a designated Google Drive root. As an initial step, we
will create and save a ggplot scatterplot as a local PNG file for the purpose
of demonstration.

```{r eval=FALSE}
library(ggplot2)
p <- ggplot(mtcars, aes(mpg, disp)) + geom_point() 
ggsave(plot = p, filename = "plots/scatterplot.png")
```

To upload this file to Google Drive, execute:

```{r eval=FALSE}
cloud_drive_upload("plots/scatterplot.png")
```

By invoking the `cloud_drive_ls()` function, you can view the automatically
created "plots" folder in the console. To inspect the contents of this folder,
which currently contains a single PNG file, use `cloud_drive_ls("plots")`. To
access the folder on Google Drive, execute `cloud_drive_browse("plots")`. To
directly view the scatterplot, use
`cloud_drive_browse("plots/scatterplot.png")`.

With `cloudfs`, you can directly write content to cloud storage, bypassing the
manual creation of local files. The file generation process remains transparent
to the user.

First, let's compute a summary of the `mtcars` dataframe:

```{r}
library(dplyr, quietly = TRUE)
summary_df <- 
  mtcars %>% 
  group_by(cyl) %>% 
  summarise(across(disp, mean))
```

To export this summary to a spreadsheet, simply specify the desired file path
with the appropriate extension. The method for writing is then inferred from
this extension:

```{r eval=FALSE}
cloud_drive_write(summary_df, "results/mtcars_summary.xlsx")
```

To view the resulting spreadsheet in Google Drive, execute
`cloud_drive_browse("results/mtcars_summary.xlsx")`.

Just as we wrote the summary to an xlsx file, we can also read from it using
`cloud_drive_read("results/mtcars_summary.xlsx")`.

It's noteworthy that the writing and reading methods are determined
automatically based on the file extension. For instance, ".xlsx" utilizes
`writexl::write_xlsx()` for reading, whereas ".csv" employs `readr::write_csv`.
A comprehensive list of default methods is available in the documentation of 
`cloud_drive_write()` and `cloud_drive_read()` functions.

Additionally, `cloudfs` offers flexibility by allowing custom writing and
reading methods. For instance, our earlier scatterplot could have been written
directly to Google Drive, bypassing local file generation:

```{r eval=FALSE}
cloud_drive_write(
  p, "plots/scatterplot.png",
  fun = \(x, file) 
    ggsave(plot = x, filename = file)
)
```

### Operations on multiple files at once

Suppose multiple CSV files have been uploaded to the "data" folder and we intend to download them locally. Instead of invoking `cloud_s3_download()` for each file, a more efficient approach is available.

But first, let's generate a few sample files for demonstration purposes.

```{r eval=FALSE}
cloud_drive_write(datasets::airquality, "data/airquality.csv")
cloud_drive_write(datasets::trees, "data/trees.csv")
cloud_drive_write(datasets::beaver1, "data/beaver1.csv")
```

Listing the contents of the "data" folder gives us the following:

```{r eval=FALSE}
cloud_drive_ls("data")
#> # A tibble: 3 × 5
#>   name           type  last_modified       size_b id      
#>   <chr>          <chr> <dttm>               <dbl> <drv_id>
#> 1 airquality.csv csv   2023-09-12 08:04:46   2890 1CXTi1A…
#> 2 beaver1.csv    csv   2023-09-12 08:04:50   1901 1Fg4s1O…
#> 3 trees.csv      csv   2023-09-12 08:04:48    400 1vDYBVt…
```

`cloudfs` offers `bulk` functions that simplify the management of multiple files
simultaneously. For instance, to download all files listed above use
`cloud_drive_download_bulk()`:

```{r eval=FALSE}
cloud_drive_ls("data") %>% 
  cloud_drive_download_bulk()
```

This action automatically downloads the datasets to a local "data" directory,
replicating the same structure as on Google Drive.

To read several CSV files from the "data" folder on Google Drive into a
consolidated list, execute:

```{r eval=FALSE}
all_data <- 
  cloud_drive_ls("data") %>% 
  cloud_drive_read_bulk()
```

To upload a collection of objects, such as ggplot visualizations, to Google
Drive, first group them in a named list. Then, utilize the `cloud_object_ls()`
function to generate a dataframe akin to the output of `cloud_drive_ls()`.
Finally, execute `cloud_drive_write_bulk()` to complete the upload.

```{r eval=FALSE}
library(ggplot2)
p1 <- ggplot(mtcars, aes(mpg, disp)) + geom_point()
p2 <- ggplot(mtcars, aes(cyl)) + geom_bar()

plots_list <- 
  list("plot_1" = p1, "plot_2" = p2) %>% 
  cloud_object_ls(path = "plots", extension = "png", suffix = "_newsletter")

plots_list %>% 
  cloud_drive_write_bulk(fun = \(x, file) ggsave(plot = x, filename = file))
```

For bulk uploads of local files to Google Drive, utilize the `cloud_local_ls()`
function. For instance, to upload all PNG files from the local "plots" directory
to Google Drive:

```{r eval=FALSE}
cloud_local_ls("plots") %>% 
  filter(type == "png") %>% 
  cloud_drive_upload_bulk()
```

### S3 functions

For Amazon S3 interactions, we offer a parallel set of functions similar to
those designed for Google Drive. These dedicated S3 functions are easily
identifiable, beginning with the prefix `cloud_s3_`.
